\chapter{Preface}

I have written this unofficial solution guide to serve as a companion
to the book {\em Number Theory}, Second Edition, by Underwood
Dudley. This manual is intended as an aid for students who are
studying number theory using Dudley's text. I strongly encourage
students using this guide to first attempt each problem for
themselves. If no progress is made after struggling with the problem
for a time, or if the student does find a solution and wants to check
their work, then this guide may be helpful.

In writing these solutions, I have avoided using any techniques or
results before the point at which they are introduced in the text. My
solutions should therefore be accessible to students who have read the
text up to the appropriate chapter.

This solution manual is lengthy and contains solutions to many
problems, so errors are inevitable. If you find an error or have a
suggestion, please feel free to email me at
\href{mailto:gkikola@gmail.com}{gkikola@gmail.com}. I appreciate any
corrections or feedback.

Please know that this guide is currently unfinished. I am slowly
working on adding the remaining chapters, but this will be done at my
own pace. If you would like to find a solution to a problem that I
have not included, try typing the problem into a web search engine
such as \href{https://www.google.com/}{Google}; it is quite likely
that someone somewhere has already solved the problem and published it
on the Internet.

This guide is licensed under the Creative Commons
Attribution-ShareAlike 4.0 International License. To view a copy of
this license, visit
\begin{center}
  \href{http://creativecommons.org/licenses/by-sa/4.0/}
  {http://creativecommons.org/licenses/by-sa/4.0/}
\end{center}

I am grateful to Underwood Dudley for authoring such a lovely and
concise reference. I also give my thanks to those readers who have
taken the time to inform me of errors in my solutions.

\begin{flushright}
  Greg Kikola\\
  \href{https://www.gregkikola.com/}{www.gregkikola.com}\\
  \href{mailto:gkikola@gmail.com}{gkikola@gmail.com}
\end{flushright}
